---
title: "Exponential Distribution Simulation"
author: "Andres Camilo Zuñiga Gonzalez"
date: "22/5/2020"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, options(scipen=999))
```

As a first step, it is necessary to load the necessary packages and set the working directory.
```{r eval=FALSE}
setwd('./Statistical_Inference')
```
```{r pacakges, message=FALSE}
library(ggplot2) #plooting system
library(cowplot) #Panel for ggplot2
```

# Exponential Distribution Simulation

## Overview
In the following exercise, I am going to show the properties of the exponential distribuiton and how to simulate this distribution and the mean of the simulation follows a normal distribution.

## Simulations
First, we define \lambda, the number of observations for each simulation and the number of simulations for the `rexp()` function.
```{r base}
lambda <- .2
n <- 40
sim <- 1000
```

Then we simulate the data and build a matrix where each row is a simulation and the columns are the observations (sim x n). The `replicate()` function performs `rexp()` the number of times defind in `sim`, while `matrix()` builds the matrix.
```{r simulation}
set.seed(941104)
simulation <- matrix(replicate(sim, rexp(n, lambda)), sim, n)
```

## Sample Mean versus Theoretical Mean
First, we must define the theoretical mean of the exponential distribution ($1/\lambda $) and find the mean for each simulation.
```{r sim_mean}
teo_mean <- 1 / lambda
sim_mean <- rowMeans(simulation)
```

Then, we build a histogram of the mean for each simulation and draw a vertical line showcasing the theoretical mean, and another line showing the simulated mean of the means.
```{r mean_hist, fig.width=10, fig.height=5, warning=FALSE, message=FALSE}
mean_hist <- ggplot() + geom_histogram(aes(x = sim_mean), color = 'gray', fill = 'lightgrey') +
     geom_vline(aes(xintercept = teo_mean), color = 'red4', size = 1) +
    geom_vline(aes(xintercept = mean(sim_mean)), color = 'blue4', size = 1) +
    labs(x = 'Simulation Mean', y = 'Frequency') +
    theme_cowplot()

mean_hist_close <- mean_hist + scale_x_continuous(limits = c(4.9, 5.1))

plot_grid(mean_hist, mean_hist_close)
```

As shown in the histograms above, where the blue line represents the simulated mean `r mean(sim_mean)`and the red show the theoretical mean `r teo_mean`, they are very close to each other. In fact, they differ by just `r teo_mean - mean(sim_mean)`.

## Sample Variance versus Theoretical Variance

## Distribution
