---
title: "Swiftkey Shiny App - Get Data"
author: "Andres Camilo Zuñiga Gonzalez"
date: "25/8/2020"
output: html_document
---
```{r setup, include=FALSE, eval=FALSE}
setwd('./Data_Science_Capstone/')
```

This is the explanation of how to get the data from the English documents for the Coursera Data Science Capstone on Word Prediction with SwiftKey Data


1. Load necessary packages

```{r packages, warning=FALSE, message=FALSE}
library(tidytext) #text handling
library(parallel) #parallel processing
```


2. Create the strings for reading the files and load the stop_words dataset from the `tidytext` package

```{r}
files <- c('en_US/en_US.blogs.txt', 'en_US/en_US.news.txt', 'en_US/en_US.twitter.txt')
types <- c('blogs', 'news', 'twitter')
data("stop_words")
stop_words_c <- stop_words$word
```

3. Since datasets are huge, I will process them in parallel using a modified `apply()` function from the `parallel` package

* Load each package and variable in the clusters with `clusterEvalQ()` and `clusterExport()`, respectively
* Sample 10% of the text lines randomly.
* Unnest tokens in n-grams and count each occurrence.
* Separate n-grams in n columns (e.g., 2-gram in two columns).
* Stop clusters.
* Rename the list.

```{r parallel-lapply, warning=FALSE, message=FALSE, results='hide'}
ncores <- 3
cl <- makePSOCKcluster(ncores)
clusterEvalQ(cl, library(readr))
clusterEvalQ(cl, library(dplyr))
clusterEvalQ(cl, library(tidyr))
clusterEvalQ(cl, library(tidytext))
clusterExport(cl, "files")
clusterExport(cl, "types")
start <- Sys.time()
word_grams <- parLapply(cl, seq(files[1:3]), function(i) {
  set.seed(123456)
  pct <- 0.1
  file <- read_lines(files[i], skip_empty_rows = T)

  text <- tibble(text = file) %>%
    sample_n(., pct * nrow(.))

  bigram_count <- text %>%
    unnest_tokens(bigram, text, token = "ngrams", n = 2) %>%
    count(bigram, sort = TRUE) %>%
    separate(bigram, c("word1", "word2"), sep = " ")
  
  trigram_count <- text %>%
    unnest_tokens(trigram, text, token = "ngrams", n = 3) %>%
    count(trigram, sort = TRUE) %>%
    separate(trigram, c("word1", "word2", "word3"), sep = " ")
  
  return(list(bigram_count = bigram_count, trigram_count = trigram_count))
})

end <- Sys.time()
time <- end - start
stopCluster(cl)

names(word_grams) <- types
```
4. Save the word column from the stop words dataset as an R object
5. Save the the list of datasets as an R object
```{r}
saveRDS(stop_words_c, file = 'SwiftkeyShinyApp/data/stop_words.rds')
saveRDS(object = word_grams, file = 'SwiftkeyShinyApp/data/word_grams.rds')
```

6. Visit the Shiny App [here](https://ancazugo.shinyapps.io/SwiftkeyShinyApp/).
7. See the pitch slides [here](https://rpubs.com/ancazugo/swiftkeyapp).