---
title: "Reproducible Research: Peer Assessment 1"
author: "Andres Camilo Zuñiga Gonzalez"
output: 
  html_document:
    keep_md: true
---
```{r eval=FALSE, echo=FALSE}
setwd('./Reproducible_Research/Week2')
```

## Loading necessary packages
```{r message=FALSE}
knitr::opts_chunk$set(options(scipen=999))
library(dplyr)
library(lattice)
```


## Loading and preprocessing the data

The dataset is loaded using `read.csv()` and the date column is formatted as a date using `as.POSIXct()`.  
```{r data-load}
activity <- read.csv('activity.csv')
activity$date <- as.POSIXct(activity$date)
```

## What is mean total number of steps taken per day?
First, a summary table must be built joining the counts by day. 
```{r byDay}
byDay <- activity %>%
    group_by(date) %>%
    summarise(TotalSteps = sum(steps))
head(byDay)
```

Then a histogram is drawn using the base plotting system on the TotalSteps column of the summary.  
```{r histogram, fig.align="center"}
#png('figures/plot1.png', width = 480, height = 480)
hist(byDay$TotalSteps, breaks = 15, main = 'Total number of steps taken per day', col = 'darkgreen', xlab = 'Steps / Day')
```

The mean number of steps taken per day is **`r mean(byDay$TotalSteps, na.rm = T)`** and the median is **`r median(byDay$TotalSteps, na.rm = T)`**.  

## What is the average daily activity pattern?
First a summary of the original dataset must be built according to the interval, just as the question above.
```{r byInterval}
byInterval <- activity %>%
    group_by(interval) %>%
    summarise(MeanSteps = mean(steps, na.rm = T))
head(byInterval)
```
Then a series plot is built using this dataFrame.  
```{r time-series, fig.align="center"}
#png('figures/plot2.png', width = 480, height = 480)
plot(MeanSteps ~ interval, byInterval, type = 'l', lwd = 2, col = 'goldenrod', main = 'Average Number of Steps per Interval', xlab = 'Interval', ylab = 'Steps Average')
```

The interval with the most steps on average is **`r byInterval[which.max(byInterval$MeanSteps),][1]`** and has **`r byInterval[which.max(byInterval$MeanSteps),][2]`** steps.  

## Imputing missing values
The number of NAs in the steps column is **`r sum(is.na(activity$steps))`**.  
For imputing NA values, first I created a copy of the original dataset. Then I iterated over the rows to check if the steps was NA and then changed its value to what the mean is for the interval of the same row. In addition, I rounded the value of the mean to 0 decimals. For this task I used the summary data for the mean per interval created above.  
```{r imputing}
activityImputed <- activity
for(n in 1:nrow(activityImputed)) {
  if(is.na(activityImputed$steps[n])) {
    new_step <- byInterval[byInterval$interval == activityImputed$interval[n],]$MeanSteps
    activityImputed$steps[[n]] <- round(new_step)
  }
}
```

For the new histogram, mean and median values per day, I created a new summary using the new dataset.
```{r new_histogram, fig.align="center"}
byDayImputed <- activityImputed %>%
    group_by(date) %>%
    summarise(TotalSteps = sum(steps))
head(byDay)

#png('figures/plot3.png', width = 480, height = 480)
hist(byDayImputed$TotalSteps, breaks = 15, main = 'Total number of steps taken per day with an Imputed Dataset', col = 'lightgreen', xlab = 'Steps / Day')
```

Now, with missing values replaced with the mean of the steps taken for that interval, the mean number of steps taken per day is **`r mean(byDayImputed$TotalSteps, na.rm = T)`** and the median is **`r median(byDayImputed$TotalSteps, na.rm = T)`**.

### Difference between imputed and not imputed

|          	| Mean     	| Median 	|
|----------	|----------	|--------	|
| With NAs 	| 10766.19 	| 10765  	|
| Imputed  	| 10765.64 	| 10762  	|


## Are there differences in activity patterns between weekdays and weekends?
For this task, I iterated over the date and checked if the day was either Saturday and Sunday (in Spanish) and assigned a value according to the condition.
```{r typeOfDay}
activityImputed$typeOfDay <- NA
for(n in 1:nrow(activityImputed)) {
  if(weekdays(activityImputed$date[n]) %in% c('sábado', 'domingo')) {
    activityImputed$typeOfDay[[n]] <- 'weekend'
  } else {
    activityImputed$typeOfDay[[n]] <- 'weekday'
  }
}
```

For the summary per interval and type of day, I grouped according to these variables. Then I created the plot using lattice.
```{r typeOfDay_plot, fig.align="center"}
byIntervalTypeOfDay <- activityImputed %>%
  group_by(interval, typeOfDay) %>%
  summarise(MeanSteps = mean(steps))
head(byIntervalTypeOfDay)

#png('figures/plot4.png', width = 480, height = 480)
xyplot(MeanSteps ~ interval | typeOfDay, byIntervalTypeOfDay, type = 'l', layout = c(1, 2), xlab = 'Interval', ylab = 'Number of Steps', main = 'Number of Steps per Type of Day and Interval')
```

